---
title: "Assignment 2: Coding Basics"
author: "Yao Yao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r echo=TRUE}
#1. 
#A sequence of numbers from one to 100, increasing by fours
sequence <- seq(1, 100, by = 4)
sequence

#2.
#Mean and median of "sequence"
mean<-mean(sequence)
median<-median(sequence)
mean
median

#3. 
#Determine whether mean is bigger than median
mean>median

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r echo=TRUE}
names <- c("Lily", "Mandy", "Jack", "Liam")
names
class(names)
#names are character

test_scores <- c(90, 89, 71, 40)
test_scores
class(test_scores)
#test_scores are numeric

pass <- test_scores>50
pass
class(pass)
#pass is logical

#Converting vectors to dataframes
df_names <- as.data.frame(names)
df_test_scores <- as.data.frame(test_scores)
df_pass <- as.data.frame(pass)

#Combining the dataframes and give names to the columns
student_test_performance <- data.frame("Names"=names,"Test Scores"=pass, "Pass or Not"=pass)
class(student_test_performance)
student_test_performance

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:  I think the main difference is that matrices only contain a single class of data, while data frames can consist of many different classes of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r echo=TRUE}
pass_test <- function(x) {
  ifelse(x>=50, TRUE, FALSE)
}
```

11. Apply your function to the vector with test scores that you created in number 5.

```{r echo=TRUE}

results <- pass_test(df_test_scores)
results
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: In my R studio, ifelse works at last, since can only check one element in a vector at one time, but using this code we attempted to check every element in the vector at once.
